Proposal: Transformer-CRF Integration for Sequence Labeling on EEG Data

### 2024-10-11 proposal/proposal.tex
proposal

\usepackage[margin = 1in]{geometry}
\usepackage{authblk}
\usepackage[colorlinks=true,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{natbib}
\usepackage{setspace}
\doublespacing

@@

\title{Proposal:
BERT-CRF Integration for Sequence Labeling on EEG Data \sx{TBD}}

@@

In recent years, deep learning has emerged as a powerful tool for sequence
labeling. Neural network architectures like long short-term memory (LSTM) and
transformer models have revolutionized sequential data processing.
Additionally, conditional random fields (CRFs) have been employed for modeling
label dependencies, demonstrating promising results.
However, while these methods have shown individual strengths, a comprehensive
approach that leverages their complementary capabilities is essential for
tackling complex, multi-modal datasets.

This project introduces a novel framework that combines deep neural networks
and probabilistic graphical models (PGMs) to enhance sequence labeling
performance. By employing bidirectional encoder representations from
transformers (BERT) for feature extraction and integrating them with CRF models,
we aim to improve sequence tagging accuracy. Furthermore, mean-field
approximation techniques are utilized for efficient inference, ensuring a
computationally efficient yet effective approach.
This integration of modern neural network architectures with CRFs offers
a robust and promising solution for complex sequence labeling tasks.

@@

Transformer-based models, such as BERT~\citep{devlin2019bert},
have emerged as powerful alternatives to traditional RNN-based approaches.
Trained on massive datasets in an unsupervised manner, BERT can be fine-tuned
for various natural language processing tasks, including sequence labeling.
\citet{devlin2019bert} directly compared BERT to traditional models, including
Bi-LSTM, and demonstrated its superior performance on tasks like named entity
recognition, highlighting its ability to capture more contextual information
than LSTM models.

@@

\citet{fan2023environment}

\section{Potential Data Sets}

This dataset consists of electroencephalography (EEG) signals collected from
hospital patients and provided by Harvard Medical School. It was originally
part of a Kaggle competition that concluded in April 2024. The data includes
spectrograms and time-series representations of EEG signals, with the goal of
classifying six patterns of harmful brain activity, such as seizure (SZ) and
generalized periodic discharges (GPD). For our project, we will segment the EEG
data into time steps and use deep neural networks for feature extraction,
followed by a fully connected CRF layer to perform the sequence labeling of
brain activity patterns.

>>>>

\usepackage[margin = 1in]{geometry}
\usepackage{authblk}
\usepackage[inline]{enumitem}
\usepackage[colorlinks=true,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{natbib}
\usepackage{setspace}
\onehalfspacing

@@

\title{Proposal: Transformer-CRF Integration for Sequence Labeling on EEG Data}

@@

In recent years, deep learning has emerged as a powerful tool for sequence
labeling. Neural network architectures like long short-term memory (LSTM) and
Transformer models have revolutionized sequential data processing.
Additionally, conditional random fields (CRFs) have been employed for modeling
label dependencies, demonstrating promising results.
However, while these methods have shown individual strengths, a comprehensive
approach that leverages their complementary capabilities is essential for
tackling complex, multi-modal datasets.

This project introduces a novel framework that combines deep neural networks
and probabilistic graphical models (PGMs) to enhance sequence labeling
performance. By employing Transformer~\citep{vaswani2017attention} for feature
extraction and integrating them with CRF models,
we aim to improve sequence tagging accuracy. Furthermore, mean-field
approximation techniques are utilized for efficient inference, ensuring a
computationally efficient yet effective approach.
This integration of modern neural network architectures with CRFs offers
a robust and promising solution for complex sequence labeling tasks.

@@

Transformer-based models, such as BERT~\citep{devlin2019bert},
have emerged as powerful alternatives to traditional RNN-based approaches.
Trained on massive datasets in an unsupervised manner, Transformer can be
fine-tuned for various natural language processing tasks,
including sequence labeling.
\citet{devlin2019bert} directly compared BERT to traditional models, including
Bi-LSTM, and demonstrated its superior performance on tasks like named entity
recognition, highlighting its ability to capture more feature information
than LSTM models.

@@

\section{Data}

The harmful brain activity classification dataset~\citep{jing2023development}
offers a valuable resource for studying brain health. This dataset,
originally used in a Kaggle competition that concluded on April 8th, 2024
(available at
\url{https://www.kaggle.com/competitions/hms-harmful-brain-activity-classification}),
contains electroencephalography (EEG) signals from hospital patients, collected
by Harvard Medical School. The data comes in two primary formats:
\begin{enumerate*}[label = (\roman*)]
\item EEG time series data, which records 50-second windows of brain activity
measured by 19 electrodes placed on the scalp;
\item Spectrogram data, which provides frequency-based representations of the
EEG signals.
\end{enumerate*}

\section{Activities Plan}

\begin{enumerate}
\item \textbf{Probability graphical model}
(Shiying Xiao and Xiaohang Ma)
\begin{itemize}
\item Build the CRF layer for label predition
\item Mean field approximation inference and training of CRF model
\end{itemize}
\item \textbf{Deep learning methods} (Shiying Xiao, Xiaohui Yin and Xiaohang Ma)
\begin{itemize}
\item CNN and RNN layers for feature extraction
\item RNN-CRF layers for label discrimination
\end{itemize}
\item \textbf{Manuscript writing} (Shiying Xiao, Xiaohui Yin and Xiaohang Ma)
\item \textbf{Project presentation} (Shiying Xiao, Xiaohui Yin and Xiaohang Ma)
\end{enumerate}
